Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-10-18 docs/manuscript/paper.tex
extra sentence for into of sections
all comments up to start of section 4

% Abstract of the paper
\begin{abstract}	
Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing array observations. A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. A two-step procedure is applied within a state-space framework, such that the model parameters are estimated with a Kalman filter, which then provides a likelihood for nested sampling. The procedure enables self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional analysis methods. It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$. Full posterior distributions of model parameters are recovered. There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms', which is analysed and discussed. Smaller biases $\lesssim 1 \%$ are also observed in some of the other parameters of the system.
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs). Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of nHz GWs has inspired the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost method is timing an ensemble of pulsars, i.e. a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}. A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth. By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA,][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}. These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep[CPTA,][]{Hobbs_2019}. \newline

@@

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs. However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}. \newline

Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations. The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density. Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization. The new and existing approaches are therefore complementary. In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}. We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood (i.e. the model evidence) for model selection. \newline

% --- a promise which we aim to test as the key goal of this paper. 
\noindent This paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative example GW system. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space. Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively. The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

\section{State-Space Formulation}\label{sec:model}
We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation. In this work we take the intrinsic state variable to be the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local,freely-falling rest frame of the pulsar's centre of mass. A phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.  We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.  The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}. The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array. The subtle problem of generalizing the measurement variable to pulse TOAs is postponed to future work, as noted in Section \ref{sec:intro}.
\subsection{Spin evolution} \label{sec:psr_frequency}
A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}. We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend. In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque. The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution $f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
\begin{equation}
 f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise. Combined with the mean reversion it gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. It is important to note that white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. \newline

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth. Specifically, the pulse frequency is modulated harmonically at the GW frequency. We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section. The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
\end{equation}
with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$. The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$. Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB). We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic. Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Zhu10},
\begin{equation}
	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_c$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. From Equation \eqref{eq:f_evolution} we can see that only those binaries which are massive or at high frequency experience significant frequency evolution over typical PTA timespans. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed systems state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are identically correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column). The top panels show the true pulsar state $f_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{1}(t) - f_{\rm em}^{(1)}(t)$  (green curve). We plot the subtracted state to better illustrate the stochastic wandering of the pulsar frequency. In the left hand column, the blue/green solutions are completely overlaid. The middle panels show the true measured frequency $f_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{1}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve). Again we plot the subtracted state, and in the left hand column the red/magenta solutions are completely overlaid. The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{1}(t) - \hat{f}_{\rm m}^{1}(t)$. The results are shown for a single pulsar. In the right hand column $\Omega$ has been perturbed from its true value by 20\%} 
	\label{fig:kalman_example}
\end{figure*}

>>>>

% Abstract of the paper
\begin{abstract}	
Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing arrays. A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. A two-step procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling. The procedure enables the static parameters in the problem, such as the sky position of the source, to be inferred without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods. It also returns the Bayes factor relating a model with a single-source gravitational wave to one without, complementing traditional detection methods. It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$. Full posterior distributions of model parameters are recovered and tested for accuracy. There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms'. Smaller biases $\lesssim 1 \%$ are also observed in some other static parameters.
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs). Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of nHz GWs has inspired the development of new observational methods, since it is impractical to engineer terrestrial interferometric detectors with sufficiently long baselines. The foremost method is timing an ensemble of pulsars, i.e. a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}. A nHz GW influences the trajectory and frequency of individual radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth. By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA,][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}. These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep[CPTA,][]{Hobbs_2019}. \newline

@@

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter, e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA}, or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A}, rather than by cross-correlating pulsar pairs. However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}. Inconclusive evidence at low significance was presented recently by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}. \newline

Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations. The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density. Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations summarized by the power spectral density, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization. The new and existing approaches are therefore complementary. In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}. We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the static GW parameters and calculate the marginal likelihood (i.e. the model evidence) for model selection. \newline

This paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space. In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates. Discussion on  future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively. The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

\section{State-Space Formulation}\label{sec:model}
We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation. In this work we take the intrinsic state variable to be the $n$-th pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass. A phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.  We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.  The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}. The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array. The subtle problem of generalizing the measurement variable to pulse TOAs is postponed to future work, as noted in Section \ref{sec:intro}.
\subsection{Spin evolution} \label{sec:psr_frequency}
A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}. We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend. In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque. The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ for the sake of definiteness \citep{1969ApJ...157..869G}. Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution $f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
 f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'} \delta (t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise. Combined with the mean reversion it gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. It is important to note that white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma^{(n)}$ in Equation \eqref{eq:frequency_evolution}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. \newline

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth. Specifically, the pulse frequency at the Earth is modulated harmonically at the GW frequency. We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section. The measurement equation is a key input into the Kalman filter in Section \ref{sec:kalman_filter}
\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
\end{equation}
with nearly Lorentz spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$ \citep{schutz2022}. The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$. Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB). We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. \newline

In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic. Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) changes during the inspiral by \citep[e.g.][]{Zhu10}
\begin{equation}
	\Delta f_{\rm gw} = 3.94 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^9 M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 10^{-7} \mathrm{~Hz}]^{11/3} (T_{\rm obs}^2/10 \mathrm{yr}) \gtrsim 8 \mathrm{yr}$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column). The top panels show the true pulsar state $f_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$  (green curve). We subtract $f_{\rm em}^{(1)}(t)$ to better illustrate the stochastic wandering of the pulsar frequency. In the left column, the blue/green solutions overlap almost perfectly; in the right column, they do not. The middle panels show the true measured frequency $f_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve). Again we subtract $f_{\rm em}^{(1)}(t)$. In the left column the red/magenta solutions overlap almost perfectly; in the right column, they do not. The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$. The results are shown for a single pulsar. In the right column $\Omega$ is displaced from its true value by 20\%.} 
	\label{fig:kalman_example}
\end{figure*}
